What We Eval in the Shadows

### 2021-04-13 paper/main.tex
origins

There are various ways to obtain expressions that \eval will evaluate:
\begin{itemize}
\item \c{substitute} synthesizes ASTs from expressions by replacing symbols
    with their bindings in the specified environment.
\item \c{match.call} captures a call and its arguments as expressions.
\item \c{quote}, \c{enquote} returns its argument unevaluated, as an R expression.
\item \c{expression} creates a vector of expression objects from text.
\item \c{parse}, \c{str2expression} and \c{str2lang} turn strings into
   expressions.
\end{itemize}
We show the distribution of the different provenances in
Table~\ref{tab:provenance}. Note that one site could be passed arguments from
different provenances depending on the invokation. For some arguments, we could
not determine their provenance or it was a value or a symbol, for which the
provenance is less significant.

\begin{table}[h]\small
\begin{tabular}{r|r|r} \hline
Origin  & \#sites & \%sites \\\hline
\c{substitute} & \packageNbOtherLangsSites & \packageOtherLangsSitePercent \\ \hline
\c{parse} & \packageNbParseExprsSites & \packageParseExprsSitePercent \\ \hline
\c{match.call} &  \packageNbMatchCallExprsSites & \packageMatchCallExprsSitePercent  \\\hline
\c{expression} & \packageNbNotParseExprsSites & \packageNotParseExprsSitePercent  \\\hline
\c{str2lang} & \packageNbStrlangSites  & \packageStrlangSitePercent  \\\hline
Other & \packageNbUnknownProvSites & \packageUnknownProvSitePercent \\\hline
\end{tabular}
\caption{Provenance}\label{tab:provenance}
\end{table}

We taint the result of a call to \c{parse}, \c{str2lang} and \c{str2expression},
by attaching an R attribute to it. R attributes are lost under copy or
modification of their attaching object, so we underestimate the number of sites
in that case. However, we observed that most of the time, the result of
\c{parse} is not touched and directly passed to \eval.

We taint the result of \c{match.call} by remembering the addresses of all the
elements of the call list. As R GC is non-moving, it makes it possible to track
them individually, as we observed that calls are modified and individual
elements are sometimes evaluated. This is also an underestimation of the number
of sites though, as we would lose an individual element of the call list that
would be modified.

% explain also how we get the numbers of the other categories?

\c{eval(parse(...))} (and \c{str2expression}, and \c{str2lang}) can be used for
dynamic code loading. This forms the core of \c{source} and \c{sys.source}
functions in R that are commonly used for loading code in R files in interactive
settings. We investigated the number of cases in which the output of \c{parse}
and its variants is passed to \eval, directly or transitively by tainting their
output. We observed that very few of the eval call (\packageNbParseFromFileSites
in total) consume the result of calling \c{parse} on a file. Most of the eval
calls consume the result of calling \c{parse} on a string. We also identified
one function in core R, \c{invokeRestartInteractively} that prompts the user for
input, parses it, and passes it to \eval.

In Kaggle, most of the sites, \ie \kaggleParseExprsSitePercent, are passed an
expressions that result from a call to \c{parse}.

>>>>

Where does the expression passed to \eval come from? There are various means of
creating that expression, these are associated with particular use cases.

\begin{itemize}
\item {\it Promise:} Function arguments are passed as promises, suspended
  computations, from a promise one can retrieve the source expression with a
  call to the \c{substitute} function.
\item {\it Reflection:} The \c{match.call} function reflectively captures
  the expression that called the current function.
\item {\it Constructed:} Expressions can be constructed by invoking the
  \c{quote}, \c{enquote} and \c{expression} functions.
\item {\it String:} Finally, expressions can be created from strings by invoking
  one of \c{parse}, \c{str2expression} and \c{str2lang}.
\end{itemize}

Table~\ref{tab:provenance} summarizes the expression provenance in our corpus.
This data is obtained by dynamically tainting values as they are produced by the
various sources, due to technical reasons there is some imprecision in the
results. Manual inspection of numerous examples suggest that ordering is correct.

\begin{table}[h]\small
\begin{tabular}{r|r|r} \hline
Origin  & \#sites & \%sites \\\hline
Promise & \packageNbOtherLangsSites & \packageOtherLangsSitePercent \\ \hline
Parse & ?& ?\\ \hline
\c{match.call} &  \packageNbMatchCallExprsSites & \packageMatchCallExprsSitePercent  \\\hline
Constructed & ? &?  \\\hline
\end{tabular}
\caption{Provenance}\label{tab:provenance}
\end{table}

The string origin can sometimes be correlated with dynamic code loading. This is
what the base functions \c{source} and \c{sys.source} do. We observed few calls
(\packageNbParseFromFileSites in total) that consume the result of calling
\c{parse} on a file. Most of the calls build strings programmatically. We also
identified one function \c{invokeRestartInteractively} that prompts the user for
input, parses it, and passes it to \eval.  The use of strings seems to correlate
with less sophisticated programmers, as in our Kaggle corpus where
\kaggleParseExprsSitePercent of site use strings.
